\section{Preliminaries: Metric and Setup}
\label{sec:prelim}

\subsection{Interventional methods for locating safety mechanisms}
Interventional methods from MI are designed to investigate the effect that the steps of the computational process in a model have on the final output. Interventional methods change model-parameters and interpret the effect of the parameters by assessing how outputs are affected by the intervention. Since it is infeasible to examine all possible combinations of parameters from modern LLMs, methods have been developed to locate the specific parameters of interest for a given behavior. We employ 3 methods that find components critical for safe behavior in LLMs: \textit{Safety Head ImPortance Score} (SHIPS) uses attention head ablation as the intervention \citep{zhou2024ships}, \textit{Safety Neuron Detection} (SND) uses neuron deactivation as the intervention \citep{zhao2025neuron}, and \textit{Refusal Direction Extraction} that uses vector steering and vector ablation as the intervention \citep{arditi2024refusal}. In the following, we will present each method as well as the interventions. First of all we have an LLM, $\mathcal{M}_{base}$ and an intervention on an LLM, $\text{intervention}(\mathcal{M}_{base}) = \mathcal{M}_{int}$, a set of potentially harmful queries, $\textbf{X}^{harm} = \{\textbf{x}_1^{harm},\textbf{x}_2^{harm},...,\textbf{x}_m^{harm}\}$ and a set of safe queries, $\textbf{X}^{safe} = \{\textbf{x}_1^{safe},\textbf{x}_2^{safe},...,\textbf{x}_m^{safe}\}$. $h(x)$ is the value in the residual stream at any given point.
\\
Attention head ablation removes the effect of an attention head, $head_i$, by multiplying the query-matrix, $W_i^Q$, with a small number, $\epsilon$:
\begin{equation*}    
    head_i^{int} = \text{softmax}(\frac{\epsilon W_i^QW_i^{KT}}{\sqrt{d/n}})W_i^V
\end{equation*}
To locate which attention heads are critical for safe behavior, we obtain $A(\mathcal{M},X_{harm})$, which is the matrix of activations from the last layer in $\mathcal{M}$, when processing $\textbf{X}_{harm}$. We then compute the SHIPS of each attention head:
\begin{equation*}
    SHIPS(A_{base},A_{int},head_i) = \sum_j^r \phi_j
\end{equation*}
where $\phi_j$ is the principal angle between the first $r$ columns of the left-singular matrix from $U_\mathcal{A_{harm}}$, and the left-singular matrix from, $U_{A_{int}}$. \textcolor{red}{not done yet...}
\\
To find a collection of heads that together has the greatest impact on safety, \citeauthor{zhou2024ships} defines the \textit{Safety Attention Head AttRibution Algorithm} (SAHARA), which takes an LLM, $\mathcal{M}$, a set of queries, $X$, a small factor, $\epsilon$ and a number of heads, $n$, and for $n$ iterations, iteratively computes the SHIPS and ablates the highest scoring head.
\\
Neuron deactivation sets all values in a neuron, $N_i$, to zero. A neuron is a row or column in the parameter matrices in an attention head or MLP. We locate neurons critical for safety from an MLP-layer, by calculating the importance of the neuron:
\begin{equation*}
    \mathcal{N}_{MLP}(x^{harm}) = ||h_{\mathrm{up}}(x^{harm}) \cdot \texttt{Mask} \bigr) W_{\mathrm{down}}||_2
\end{equation*}
where $h_{\mathrm{up}}(x^{harm})$ is the activation after up-projection, \texttt{Mask} is the identity-matrix of size $\text{dim}(h_{\mathrm{up}}) \times \text{dim}(h_{\mathrm{up}})$ and $W_{\mathrm{down}}$ is the down-projection matrix.
\\
Calculating the importance of neurons from attention head layers, we use the formula:
\[\resizebox{\columnwidth}{!}{$\displaystyle
\begin{aligned}
    \mathcal{N}_{att}(x^{harm}) = &\left\|\operatorname{softmax}\!\left(
\frac{W^Q(x^{harm})W^{kT}(x^{harm})-\Delta(x^{harm})}{\sqrt{d}}
\right)\right. \\
&\left.-
\operatorname{softmax}\!\left(
\frac{W^Q(x^{harm})W^{KT}(x^{harm})}{\sqrt{d}}
\right)\right\|_2
\end{aligned}
$}\]
where
\[\resizebox{\columnwidth}{!}{$\displaystyle
    \Delta(x^{harm}) = W^Q(x^{harm}).\operatorname{resize}(l,1,d)
\times
W^K(x^{harm}).\operatorname{resize}(1,l,d)
$}\]
The collection of neurons important for safety satisfies: 
\begin{equation*}
    \mathcal{N}_{\mathrm{safe}} = \{\, N_i \mid N_i \geq \epsilon,\ \forall \textbf{x}^{harm} \in \textbf{X}^{harm} \,\}
\end{equation*}
\\
Vector steering takes a vector, $\textbf{r}$, that represents a given behavior and adds it to the residual stream at a given layer, $l$:
\begin{equation*}
    h(x)^{(l)}_{steer} = h(x)^{(l)} + \textbf{r}^{(l)}
\end{equation*}
Vector ablation takes a unit-norm vector, $\hat{\textbf{r}}$, that represents a given behavior and removes the direction from the residual stream in every layer in the network:
\begin{equation*}
    h(x)_{abl} = h(x) - \hat{\textbf{r}}\hat{\textbf{r}}^Th(x)
\end{equation*}
To locate the vector that represents refusing harmful queries, for each layer, $l$, at each token position, $i$, we calculate the mean activations for harmful queries, $\boldsymbol{\mu}^{(l)}$, and for safe queries, $\boldsymbol{\nu}^{(l)}$, and find the difference:
\begin{align*}
    \boldsymbol{\mu}^{(l)}_i &= \frac{1}{|\textbf{X}^{harm}|} \sum_{\textbf{X}^{harm}}h(x^{harm}_i)^{(l)} \\
    \boldsymbol{\nu}^{(l)}_i &= \frac{1}{|\textbf{X}^{safe}|} \sum_{\textbf{X}^{safe}}h(x^{safe}_i)^{(l)} \\
    \textbf{r}^{(l)}_i &=  \boldsymbol{\mu}^{(l)}_i - \boldsymbol{\nu}^{(l)}_i
\end{align*}
This gives $I \times L$ candidate vectors, and we choose the best one by validating their ability to bypass refusal.

\subsection{What would finding a safety mechanism entail?}
\label{sec:entailments}
The interventions we evaluate are advanced as evidence that safety has been
\emph{located}: in particular attention heads \citep{zhou2024ships}, in sparse MLP
neurons, or in a single linear direction of the residual stream
\citep{arditi2024refusal,zou2023representation}. We take no position on which of these
objects---if any---deserves the name ``safety mechanism.'' The term has no agreed
referent, and the claims are not stated in a common vocabulary in which one could be
tested against another. What \emph{can} be tested is a claim's behavioural consequence.
Whatever a safety mechanism is, removing one entails three things:
\begin{enumerate}[leftmargin=1.4em, itemsep=1pt, topsep=2pt]
  \item \textbf{Necessity.} Safety behaviour must actually degrade. A candidate whose
  removal changes nothing was not load-bearing for safety.
  \item \textbf{Selectivity.} \emph{Only} safety behaviour may degrade. If general
  coherence collapses too, the intervention hit something broader than safety.
  \item \textbf{Depth.} The change must be visible in the safety
  \emph{computation}---the model's own reasoning about whether to comply---not merely
  in the surface form of the final answer.
\end{enumerate}
The three axes below operationalise exactly these entailments: Potency measures
necessity, Quality measures selectivity, and Safety-Reasoning measures depth. A method
scores highly only when it behaves \emph{as if} it found a safety mechanism. The score
therefore grades the \emph{localization claim}, not the ontology behind it; we return
to what the signature can and cannot establish in \S\ref{sec:discussion} and the
Limitations.

\subsection{Three axes}
\label{sec:axes}
We score an intervention on a \emph{cell} $c$---a \texttt{(model, dataset, condition)}
triple---along three axes that answer three separate questions: did it remove safety, did
it keep the model intact, and did the visible reasoning still engage safety? Each axis is
mapped to $[0,1]$ and, crucially, \emph{baseline-corrected} against the same-\allowbreak
\texttt{(model, dataset)} baseline cell $b$ (the model with no intervention). Writing
$\clip(x)=\max(0,\min(1,x))$,
\begin{align}
  \Ppot   &= \clip\!\left[\tfrac{\HAC_c-\HAC_b}{1-\HAC_b}\right],
  & \Qual  &= \clip\!\left[\tfrac{\mathrm{clean}_c}{\mathrm{clean}_b}\right],
  \label{eq:pq}\\[2pt]
  \Sreason &= 1-\clip\!\left[\,|\gapm_c|-|\gapm_b|\,\right]. &&
  \label{eq:s}
\end{align}

\paragraph{Potency ($\Ppot$).} $\HAC$ is \emph{harmful-among-clean}: the attack-success
rate computed \emph{only} over responses that pass a coherence gate. Gating first is what
defeats the ``broke the model, didn't make it harmful'' loophole---incoherent output
cannot count as a successful attack. $\Ppot$ is then the share of the
still-refusable headroom $(1-\HAC_b)$ that the intervention converts into coherent harm.

\paragraph{Quality ($\Qual$).} $\mathrm{clean}$ is the coherence-retention rate. $\Qual$
falls below $1$ exactly when an intervention degrades the model rather than selectively
removing its safety. (Quality is coherence-only here; benign-utility retention is future
work, \S\ref{sec:limitations}.)

\paragraph{Safety-Reasoning ($\Sreason$).} The CoT-monitorability gap
$\gapm=\mathrm{asr}_{\text{final}}-\mathrm{asr}_{\text{cot-pred}}$ compares the harm of the
full answer with what a judge predicts from the \emph{reasoning trace alone}. A large
positive gap means unsafe answers behind benign-looking traces (monitorability collapsed,
the worst case); $\Sreason$ rewards keeping the gap near the baseline's.

Baseline-correction is the load-bearing choice: it isolates the effect the intervention
\emph{induces} from behaviour the base model already had, so the axes remain comparable
across models whose base safety differs. We adopt a \emph{suppressive} orientation---a
high score is a potent, coherence-preserving, still-monitorable removal of answer-safety;
a defence evaluation simply flips the sign of $\Ppot$.

\subsection{The composite score}
\label{sec:composite}
We report the $(\Ppot,\Qual,\Sreason)$ vector with a Pareto front as the primary object,
and one headline scalar for the reader who wants a single number---the
\textbf{Selective-Failure Score}
\begin{equation}
  \SFS = \left(\Ppot\cdot\Qual\cdot\Sreason\right)^{1/3}.
  \label{eq:sfs}
\end{equation}
The geometric mean gives a ``no axis left behind'' property: any axis going to zero
collapses the score, so a method cannot buy a high $\SFS$ by trading a broken model for
raw potency. Reporting the scalar always \emph{with} the vector follows HELM's
multi-metric precedent \citep{liang2022helm}. Two variants appear in the ablation
(\S\ref{sec:appendix-metric}): $\SFS_{\text{prod}}=\Ppot\Qual\Sreason$ and the threat-oriented
$\SFS_{\text{covert}}=\Ppot\Qual(1-\Sreason)$. Figure~\ref{fig:schematic} summarises.

\begin{figure}[t]
  \centering
  \resizebox{\columnwidth}{!}{%
  \begin{tikzpicture}[
    font=\small,
    axisbox/.style={rounded corners=2.5pt, draw=black!55, line width=0.6pt,
                    inner sep=4pt, align=left, text width=3.15cm, fill=black!3},
    op/.style={circle, draw=black!60, line width=0.7pt, inner sep=1.5pt,
               align=center, fill=black!6, minimum size=1.05cm},
    out/.style={rounded corners=2.5pt, draw=black!70, line width=0.9pt,
                inner sep=4pt, align=center, fill=black!8},
    arr/.style={-{Stealth[length=4pt]}, draw=black!55, line width=0.7pt}]
    \node[axisbox] (P) {\textbf{Potency $P$}\\[1pt]
      {\scriptsize removed safety?}\\ {\scriptsize induced coherent harm}};
    \node[axisbox, below=5pt of P] (Q) {\textbf{Quality $Q$}\\[1pt]
      {\scriptsize model intact?}\\ {\scriptsize coherence retention}};
    \node[axisbox, below=5pt of Q] (S) {\textbf{Safety-Reasoning $S$}\\[1pt]
      {\scriptsize CoT still safe-aware?}\\ {\scriptsize monitorability}};
    \node[op, right=14pt of Q] (gm) {geo.\\mean};
    \node[out, right=14pt of gm] (sfs) {$\mathrm{SFS}$\\[1pt]
      {\scriptsize $(PQS)^{1/3}$}};
    \draw[arr] (P.east) -- (gm);
    \draw[arr] (Q.east) -- (gm);
    \draw[arr] (S.east) -- (gm);
    \draw[arr] (gm) -- (sfs);
  \end{tikzpicture}}
  \caption{\textbf{The metric.} Three baseline-corrected axes (Eqs.~\ref{eq:pq}--\ref{eq:s})
    are reported as a $(P,Q,S)$ vector with a Pareto front, and combined by geometric
    mean into the headline Selective-Failure Score (Eq.~\ref{eq:sfs}). Any axis $\to 0$
    collapses the score.}
  \label{fig:schematic}
\end{figure}

\subsection{Experimental grid}
\label{sec:grid}
We evaluate on one controlled grid of 5 models $\times$ 2 datasets $\times$ up to 11
conditions (Table~\ref{tab:grid}), with iso-ASR anchoring and layer-matched-random
controls so that families are compared at matched strength rather than at each paper's
preferred operating point. The four intervention families instantiate the distinct
mechanistic targets a metric must be able to separate---per-head, per-neuron,
dose-controlled activation-addition, and full directional ablation.

\begin{table}[t]
  \centering \small
  \begin{tabular}{@{}ll@{}}
    \toprule
    \textbf{Axis} & \textbf{Values} \\
    \midrule
    Models & \texttt{qwen3\_8b}, \texttt{olmo3\_7b\_think}, \\
           & \texttt{olmo3\_7b\_base}, \texttt{olmo3\_7b\_base\_own}, \\
           & \texttt{llama31\_8b\_control} \\
    \addlinespace[2pt]
    Datasets & JailbreakBench (100), BeaverTails (98) \\
    \addlinespace[2pt]
    Conditions & \texttt{baseline}; \texttt{ships\_top\{3,5,8\}} (heads); \\
               & \texttt{neurons\_top\{256,512,1024\}}; \\
               & \texttt{steering\_a\{0.5,1.0,1.5\}} (add); \\
               & \texttt{steering\_ablate} (dir.\ ablation) \\
    \bottomrule
  \end{tabular}
  \caption{The evaluation grid. \texttt{steering\_a1.0} is the iso-ASR anchor.}
  \label{tab:grid}
\end{table}

\subsection{Judges and their validation}
\label{sec:judges}
Three LLM judges instrument the axes: a 5-label safety + coherence judge (Qwen3-30B)
feeds $\Ppot$ and $\Qual$; a per-sentence safety-reasoning trace judge and a CoT-only
predictor (Qwen3-30B, vLLM backend) feed $\Sreason$; and a fine-tuned 12-label pathway
judge (Qwen3-14B LoRA) provides the mechanistic decomposition behind $\Sreason$. A metric
is only standardizable if its instruments are reliable: the pathway judge reaches
$\kappa\approx0.96$ / $F_1\,0.98$ against a human gold set (a 30B baseline scores
$\kappa\,0.21$), and the 5-label, CoT-only, and trace judges are human-validated on the
\texttt{batch\_v5\_002} annotation task by two annotators. \TODO{full $\kappa$ table
(Table 2) + inter-annotator agreement.}
