% =========================================================================
% 5. DISCUSSION   [~0.5 page]
% GOAL: zoom out from "our grid" to "why the instrument matters." No new results.
% =========================================================================
\section{Discussion}
\label{sec:discussion}

\paragraph{The measurement problem is not specific to the four intervention methods.}
The sharpest practical finding of \S\ref{sec:compare} is that a raw attack success rate re-ranks interventions and over credits already unsafe models. Any white-box intervention evaluated with an absolute rate inherits its model's baseline harm, and the cheap correction provided with our potency $\Ppot$ metric alleviates this issue. We encourage intervention papers to report baseline-corrected potency alongside a coherence or quality axis by default, independent of whether the composite metric is adopted. Together, this makes intervention results across models and papers comparable. 

\paragraph{Most localizations do not localize behaviorally.}
At matched strength, the direction-based interventions leave the signature of a removed safety mechanism, i.e., they induce coherent harm without any loss of coherence. The safety head and neuron interventions are not indicative of this; removing the components they identify changes little. Our falsification claim supports this: a discovery of safety artifacts in a model with no safety training returns confident targets, the removal of which induces effectively no harm. We are correspondingly careful about what a high score licenses. A potent, selective, monitorable intervention behaves as if it found a safety mechanism, necessary but not sufficient.

\paragraph{Where the safety computation appears to sit.} 
The chain-of-thought results (\S\ref{sec:cot}) indicate that across families, the intervention leaves recognition and refusal initiation near baseline while raising execution and operational detail; the labels that move with induced harm are the downstream ones, and explicit refusal-suppression barely moves. These interventions therefore override the safety mechanism's effect on the \emph{output} rather than deleting the upstream computation that recognizes harm and begins to refuse. That the direction methods succeed where component-ablation fails fits the same picture: safety behavior here looks more like a recoverable direction acted on late than a small set of components whose removal erases it.

\paragraph{Good news for chain-of-thought monitoring, with an expiry condition.} The
monitorability result is, for once, reassuring: none of these interventions buys
potency by hiding harm from the trace. Even at the strongest steering dose, the
reasoning stays at least as revealing as the answer---more conflicted, not more
concealing---so an auditor reading the chain-of-thought would still catch the harm.
This is direct evidence for the premise of the CoT-monitoring agenda
\citep{baker2025cotmonitoring, korbak2025monitorability}: against this class of
white-box suppression, monitoring the reasoning is robust. The caveat is that the
guaranty belongs to models that \emph{externalize} their reasoning---it held on the
two explicitly-reasoning models and had little to grip on the others, whose traces
rarely verbalize safety. As models move toward shorter or more latent reasoning, the
monitorable surface these interventions leave intact may cease to exist, and the
positive result would not transfer. Monitorability is worth measuring precisely
because it is contingent.
 
\paragraph{Limitations.}
\label{sec:limitations}
Several caveats bound these claims. Our selectivity axis
measures coherence rather than capability, so an intervention that remains fluent while
undermining task competence would score better than it deserves; extending $\Qual$
with a benign-utility delta is the most important open improvement to the metric. The
axes rest on LLM judges validated against human annotation, with moderate agreement
on the labels driving the headline axes and near-chance agreement on two auxiliary
and three fine-grained labels that we therefore exclude from every axis; in
particular, the monitorability result depends on a CoT-only predictor with moderate
agreement, so it is best read as no covert failure \emph{detectable at this
reliability}. The grid consists of five 7--8B dense, text-only models evaluated at a single
seed with inference-time interventions only: the axes are scale-free by construction,
but the family ordering is not yet verified at a larger scale, on
mixture-of-experts or multi-modal models, or against fine-tuning-based methods, and
our intervals are bootstrapped over prompts rather than over generations. Finally, the
two pre-alignment \model{OLMo-3-base} arms are controls on whether a mechanism exists
to find one, not a faithfulness comparison between methods.
 
\paragraph{Future work.} The most direct next test of a localization claim follows
from the metric being orientation-agnostic: a genuine safety direction should be
drivable \emph{both} ways, so an intervention that weakens safety when removed should
harden it when added, an experiment that the same three axes score by flipping the sign of
$\Ppot$. Three further directions matter. Whether the monitorability guarantee
survives models with compressed or latent reasoning is, by the argument above, the
question that decides how far our positive CoT result generalizes. The quality axis
should be extended from coherence to benign-utility retention, so that selectivity is
measured against capability and not only fluency. Model resilience, which refers to how much harm
the most effective known intervention extracts, varies substantially across our
models and is not predicted by baseline safety. It is worth tracking as a comparable
alignment-robustness property as training recipes change.

% ---- ¶1 (What the metric buys future work): decomposing "removes safety" into P,Q,S
%      and baseline-correcting each gives a comparable, reusable number that survives
%      porting across models whose base safety differs. The (P,Q,S) vector + Pareto
%      front keeps the decision transparent; the scalar answers the one-number reviewer.

% ---- ¶2 (Mechanism as the "why" behind S): the 12-label pathway taxonomy
%      (recognition / refusal-dynamics / rationalisation / execution) is reported as a
%      descriptive dominant_pathway histogram, explaining HOW safety reasoning changes
%      on a cell without contaminating a comparable scalar. One worked example.

% ---- ¶2b (What the grid says about the localization claims): read the findings back
%      through the entailments frame (§3.1). At matched strength, the direction-based
%      hypotheses pass the signature (necessity + selectivity + depth); head/neuron
%      localizations mostly fail necessity, and the base_own control shows discovery
%      returns confident candidates even where there is no mechanism to find.
%      Consistent with safety being low-rank and distributed rather than a small
%      circuit. State strictly as evidence about the CLAIMS at these budgets --- not
%      "safety is a direction".

% ---- ¶2c (Epistemic humility --- necessary, not sufficient): the behavioural
%      signature is necessary evidence of mechanism identification, never sufficient.
%      High P shows the target is causally sufficient to DISRUPT safety, not that it
%      IS the mechanism --- downstream or correlated substrates can score identically,
%      and interventions with identical signatures form an equivalence class the
%      metric cannot break. The metric grades localization claims on their own
%      operational terms; certifying ontology needs circuit-level causal analysis
%      (complementary, not replaced). This pre-empts the mech-interp objection.

% ---- ¶3 (From suppression to defence): the metric is orientation-agnostic --- a
%      defence eval flips the sign of P and reuses Q, S unchanged, so the same
%      instrument scores both breaking and hardening a model.

% ---- ¶4 (Scope / what generalises): 7--8B here; the axes and baseline-correction are
%      scale-free by construction; generalising beyond suppressive interventions and
%      beyond 8B is future work (forward-ref Limitations).
